\begin{thebibliography}{99}

\bibitem{mirkovic2004taxonomy}
J. Mirkovi\'{c} and P. Reiher, ``A taxonomy of DDoS attack and DDoS defense mechanisms,'' \textit{ACM SIGCOMM Computer Communication Review}, vol. 34, no. 2, pp. 39--53, 2004.

\bibitem{antonakakis2017understanding}
M. Antonakakis \textit{et al.}, ``Understanding the Mirai botnet,'' in \textit{Proceedings of the 26th USENIX Security Symposium}, pp. 1093--1110, 2017.

\bibitem{silva2013botnets}
S. S. C. Silva, R. M. P. Silva, R. C. G. Pinto, and R. M. Salles, ``Botnets: A survey,'' \textit{Computer Networks}, vol. 57, no. 2, pp. 378--403, 2013.

\bibitem{beigi2014towards}
E. B. Beigi, H. H. Jazi, N. Stakhanova, and A. A. Ghorbani, ``Towards effective feature selection in machine learning-based botnet detection approaches,'' in \textit{IEEE Conference on Communications and Network Security (CNS)}, pp. 247--255, 2014.

\bibitem{chen2016xgboost}
T. Chen and C. Guestrin, ``XGBoost: A scalable tree boosting system,'' in \textit{Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining}, pp. 785--794, 2016.

\bibitem{ke2017lightgbm}
G. Ke \textit{et al.}, ``LightGBM: A highly efficient gradient boosting decision tree,'' in \textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 30, pp. 3146--3154, 2017.

\bibitem{mirsky2018kitsune}
Y. Mirsky, T. Doitshman, Y. Elovici, and A. Shabtai, ``Kitsune: An ensemble of autoencoders for online network intrusion detection,'' in \textit{Network and Distributed System Security Symposium (NDSS)}, 2018.

\bibitem{gu2008botminer}
G. Gu, R. Perdisci, J. Zhang, and W. Lee, ``BotMiner: Clustering analysis of network traffic for protocol-and structure-independent botnet detection,'' in \textit{Proceedings of the 17th USENIX Security Symposium}, pp. 139--154, 2008.

\bibitem{sharafaldin2018toward}
I. Sharafaldin, A. H. Lashkari, and A. A. Ghorbani, ``Toward generating a new intrusion detection dataset and intrusion traffic characterization,'' in \textit{Proceedings of the 4th International Conference on Information Systems Security and Privacy (ICISSP)}, pp. 108--116, 2018.

\bibitem{lundberg2017unified}
S. M. Lundberg and S.-I. Lee, ``A unified approach to interpreting model predictions,'' in \textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol. 30, pp. 4765--4774, 2017.

\bibitem{ribeiro2016should}
M. T. Ribeiro, S. Singh, and C. Guestrin, ``Why should I trust you? Explaining the predictions of any classifier,'' in \textit{Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining}, pp. 1135--1144, 2016.

\bibitem{chawla2002smote}
N. V. Chawla, K. W. Bowyer, L. O. Hall, and W. P. Kegelmeyer, ``SMOTE: Synthetic Minority Over-sampling Technique,'' \textit{Journal of Artificial Intelligence Research}, vol. 16, pp. 321--357, 2002.

\bibitem{scikit-learn}
F. Pedregosa \textit{et al.}, ``Scikit-learn: Machine learning in Python,'' \textit{Journal of Machine Learning Research}, vol. 12, pp. 2825--2830, 2011.

\bibitem{fastapi}
S. Ram\'{i}rez, ``FastAPI: A modern, fast web framework for building APIs with Python 3.7+,'' Available: \url{https://fastapi.tiangolo.com/}, 2019.

\bibitem{streamlit}
Streamlit Inc., ``Streamlit: The fastest way to build and share data apps,'' Available: \url{https://streamlit.io/}, 2019.

\end{thebibliography}
